---
title: "Cyclistic Case Study"
author: "Miro Snejdr"
date: "2022-09-10"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

I am a junior data analyst working in the marketing analyst team at
Cyclistic, a bike-share company in Chicago. The director of marketing
believes the company's future success depends on maximizing the number
of annual memberships. Therefore, my team wants to understand how casual
riders and annual members use Cyclistic bikes differently. From these
insights, my team will design a new marketing strategy to convert casual
riders into annual members. But first, Cyclistic executives must approve
my recommendations, so they must be backed up with compelling data
insights and professional data visualizations.

## Data analysis steps for this project:

-   Ask
-   Prepare
-   Process
-   Analyze
-   Share
-   Act

### Ask

1\. How do annual members and casual riders use Cyclistic bikes
differently?\
2. Why would casual riders buy Cyclistic annual memberships?\
3. How can Cyclistic use digital media to influence casual riders to
become members?

**Business task**

The main objective is to build the best marketing strategies to turn
casual bike riders into annual members by analyzing how the 'Casual' and
'Annual' customers use Cyclistic bike share differently.

### Prepare

I will use Cyclistic's historical trip data to analyze and identify
trends. The data has been made available by Motivate International Inc.
under this [license](https://www.divvybikes.com/data-license-agreement).
Datasets are available
[here](https://divvy-tripdata.s3.amazonaws.com/index.html).\

-   Download data: I'm going to use data for the last 12 months.
-   Identify: each month is a separate .csv file.
-   Determine credibility: these datasets are appropriate and will
    enable to answer the business questions.

**Installing and loading necessary packages**

```{r}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
install.packages("janitor", repos = "http://cran.us.r-project.org")
install.packages("ggplot2", repos = "http://cran.us.r-project.org")
install.packages("geosphere", repos = "http://cran.us.r-project.org")
install.packages("lubridate", repos = "http://cran.us.r-project.org")
library(tidyverse)
library(janitor)
library(ggmap)
library(geosphere)
library(lubridate)
```

**Importing our data in R studio**\
<!-- we create a data frame for each month and use the read_csv function -->

```{r}
sep21 <- read_csv("202109-divvy-tripdata.csv")
oct21 <- read_csv("202110-divvy-tripdata.csv")
nov21 <- read_csv("202111-divvy-tripdata.csv")
dec21 <- read_csv("202112-divvy-tripdata.csv")
jan22 <- read_csv("202201-divvy-tripdata.csv")
feb22 <- read_csv("202202-divvy-tripdata.csv")
mar22 <- read_csv("202203-divvy-tripdata.csv")
apr22 <- read_csv("202204-divvy-tripdata.csv")
may22 <- read_csv("202205-divvy-tripdata.csv")
jun22 <- read_csv("202206-divvy-tripdata.csv")
jul22 <- read_csv("202207-divvy-tripdata.csv")
aug22 <- read_csv("202208-divvy-tripdata.csv")
```

**Let's check the data to see if the column names and the data
structures are consistent**\
<!-- we use the colnames() and str() functions -->

```{r}
colnames(sep21)
colnames(oct21)
colnames(nov21)
colnames(dec21)
colnames(jan22)
colnames(feb22)
colnames(mar22)
colnames(apr22)
colnames(may22)
colnames(jun22)
colnames(jul22)
colnames(aug22)
```

```{r}
str(sep21)
str(oct21)
str(nov21)
str(dec21)
str(jan22)
str(feb22)
str(mar22)
str(apr22)
str(may22)
str(jun22)
str(jul22)
str(aug22)
```

### Process

**We can now merge all these montly data frames into a single large data
frame called tripdata**\
<!-- we use the bind_rows() function -->

```{r}
tripdata <- bind_rows(sep21, oct21, nov21, dec21, jan22, feb22, mar22, apr22, may22, jun22, jul22, aug22)
```

**Let's check our new dataframe**\
<!-- we use the following functions: head(), tail(), str(), summary() -->

```{r}
head(tripdata)
```

```{r}
str(tripdata)
```

```{r}
summary(tripdata)
```

**Let's add some new columns by extracting date, month, year, day of the
week into separate columns. We will use the started_at and ended_at
columns for the extraction.**\
<!-- we use the mutate() function -->\
<!-- https://mgimond.github.io/ES218/Week02c.html -->

```{r}
tripdata <- tripdata %>% 
mutate(year = format(as.Date(started_at), "%Y")) %>% # extract year, new column name is year
mutate(month = format(as.Date(started_at), "%B")) %>% #extract month, new column name is month
mutate(date = format(as.Date(started_at), "%d")) %>% # extract date, new column name is date
mutate(day_of_week = format(as.Date(started_at), "%A")) %>% # extract day of week, new column name is day_of_week
mutate(ride_length = difftime(ended_at, started_at)) %>% #extract ride length, new column name is ride_length
mutate(start_time = strftime(started_at, "%H")) #extract start time, new column name is start_time

#another way of creating a new column for year:
#trip_data$year <- format(as.Date(trip_data$date), "%Y")
#another way of creating a new column for ride_length:
#trip_data$ride_length <- difftime(trip_data$ended_at, trip_data$started_at)
```

**Now we convert ride_length into the numeric type and check it. The
value is in seconds.**

```{r}
tripdata <- tripdata %>% 
  mutate(ride_length = as.numeric(ride_length))
```

```{r}
is.numeric(tripdata$ride_length)
```

**Let's add the ride distance in km**\
<!-- calculating the distance between start_lng, start_lan and end_lng, end_lat -->
<!-- https://rdrr.io/cran/geosphere/man/distGeo.html -->
<!-- https://www.statology.org/r-add-a-column-to-dataframe/ -->

```{r}
tripdata$ride_distance <- distGeo(matrix(c(tripdata$start_lng, tripdata$start_lat), ncol = 2), matrix(c(tripdata$end_lng, tripdata$end_lat), ncol = 2))
```

<!-- now we convert this value to km -->

```{r}
tripdata$ride_distance <- tripdata$ride_distance/1000
```

**We remove rows where the ride_length has a zero or negative value and
name this cleaned data frame as tripdata_clean**\

```{r}
tripdata_clean <- tripdata[!(tripdata$ride_length <= 0),]
```

<!-- let's check our new dataframe -->

```{r}
str(tripdata_clean)
```

<!-- and a summary -->

```{r}
summary(tripdata_clean)
```

### Analyze

We are now interested in an average ride length, longest and shortest
rides.\

```{r}
tripdata_clean %>% 
  summarise(average_ride_length = mean(ride_length),max_ride_length = max(ride_length), min_ride_length = min(ride_length))
```

<!-- How to display the minimal value for a particular column: min(dataframe$column) -->

**Let's calculate the difference in total rides taken between members
and casual users. The column member_casual contains the values member
and casual.**

```{r}
tripdata_clean %>% 
    group_by(member_casual) %>% #groups by the member_casual column
    summarise(ride_count = n()) #counts all rows win ride_count column
              
```

```{r}
#we can also use the length function, here it calculates the length of the entire column ride_id
#https://www.statology.org/length-function-in-r/

tripdata_clean %>% 
    group_by(member_casual) %>% #adds a member_casual column
    summarise(ride_count=length(ride_id), ride_percentage = (length(ride_id) / nrow(tripdata_clean))*100)
```

**We can display the ratio of casuals vs members in number of rides**

```{r}
ggplot(tripdata_clean, aes(x = member_casual, fill=member_casual)) +
    geom_bar() +
    labs(x="Casuals vs Members", y="Number Of Rides", title= "Casuals vs Members Distribution") 

```

```{r}
ggplot(tripdata_clean, aes(x = rideable_type, fill=member_casual)) +
    geom_bar() +
    labs(x="Rideable Type", y="Number of Rides", title= "Total Rides by Casuals and Members by Rideable Type") 

```

**Now let's do more calculations. Values will be displayed in minutes
because we are dividing them by 60.**

```{r}
tripdata_clean %>%
  group_by(member_casual) %>% 
  summarise(average_ride_length = mean(ride_length) / 60, max_ride_length = max(ride_length) / 60, min_ride_length = min(ride_length) / 60)
```

### Conclusion 1

Casuals are taking longer rides and fewer rides than members.\

**We want to order the days from Monday to Sunday**

<!-- tripdata_clean$day_of_week <- ordered(tripdata_clean$day_of_week,  -->

<!--                                     levels=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")) -->

```{r}
tripdata_clean %>% 
  group_by(member_casual, day_of_week) %>%  #groups by member_casual
  summarise(number_of_rides = n() #calculates the number of rides and average duration 
  ,average_ride_length = mean(ride_length),.groups="drop") %>% # calculates the average duration
  arrange(member_casual, day_of_week) #sort
```

**We can now display total rides in a day of the week for both casuals
and members:**

```{r}
tripdata_clean %>%  

 group_by(member_casual, day_of_week) %>% 

 summarise(number_of_rides = n(), .groups="drop") %>% 

 arrange(member_casual, day_of_week) %>% 

 ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +

 labs(title ="Total rides of Members and Casual riders ss Day of the week") +

 geom_col(width=0.5, position = position_dodge(width=0.5)) +

 scale_y_continuous(labels = function(x) format(x, scientific = FALSE))

```

**Let's look at number of bike rides per week**\

```{r}
tripdata_clean %>%
  group_by(member_casual,day_of_week) %>%
  dplyr::summarize(count_trips = n()) %>%  
  ggplot(aes(x= day_of_week, y=count_trips, fill=member_casual, color=member_casual)) +
    geom_bar(stat='identity', position = 'dodge') +
    theme_bw()+
    labs(title ="Number of bike rides per Week", x = "Day", y = "Count of Trips")
```

**And here is the number of bike rides per month**

```{r}
tripdata_clean %>%
  group_by(member_casual,month) %>%
  dplyr::summarize(count_trips = n()) %>%  
  ggplot(aes(x= month, y=count_trips, fill=member_casual, color=member_casual)) +
    geom_bar(stat='identity', position = 'dodge') +
    theme_dark() +
    labs(title ="Number of bike rides per month", x = "Month", y = "Count of Trips")
```

**What are the 10 most popular starting stations for casuals?**

```{r}
tripdata_clean %>%
    group_by(member_casual,start_station_name) %>%
    dplyr::summarise(number_of_ride = n()) %>%
    filter(start_station_name !="", "casual"==member_casual) %>%
    arrange(-number_of_ride) %>%
    head(n=10) %>%
    select(-member_casual)
```

**What are the 10 most popular starting stations for members?**

```{r}
tripdata_clean %>%
group_by(member_casual,start_station_name) %>%
dplyr::summarise(number_of_ride = n()) %>%
filter(start_station_name !="", "member"==member_casual) %>%
arrange(-number_of_ride) %>%
head(n=10) %>%
select(-member_casual)
```

# Summary

Learned that a docked bicycle type is around 2 times as popular compared
to casual and electric bikes. This applies to both casual and member
bicycle users.

Learned that Saturday and Sunday are the most popular riding days for
casual riders.

November through February have the least number of casual riders,
perhaps due to cold winter months.

May, July, and August have a particularly high number of Casual riders.

The most popular stations for Casual riders in descending order are
Streeter Dr & Grand Ave, Lake Shore Dr & Monroe St, Millennium Park,
Theater on the Lake, Michigan Ave & Oak St.

Conclusion: Based on the data analyzed I would recommend we focus our
marketing effors for Casual Riders with these parameters 1, Increase
marketing for docket bicycles 2. Heavier marketing from May through
August 3. Focus marketing on weekends 4. Invest in marketing at the top
5 stations as noted above. Top 10 station marketing would cover even
more ground.
