---
title: "Rtutorial"
author: "Miro Snejdr"
date: "2022-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basics

How to get help?

```{r}
?function
```

You can use R like a calculator

```{r}
5+6
```

We can assign a value to an object:

```{r}
a <- 5
b <- 6

a+b #first way of counting these values
sum(a+b) #another way, sum is a function
```

Creating variables with values. Ages is the variable:

```{r}
ages <-c(48, 62, 48, 87, 23)
```

Let's check the variable

```{r}
ages
```

We can sum (for example) all these values just by using the sum function and the name of the variable

```{r}
sum(ages)
```

Values can be text as well. Let's create another variable names:

```{r}
names <- c("richard", "john", "james", "edward", "donald")
```

And check it

```{r}
names
```

Let's create a data frame called friends by using the data.frame function:

```{r}
friends <- data.frame(names, ages)
```

Check:

```{r}
friends
```

```{r}
View(friends) #shows the data frame in a table
```

```{r}
str(friends) #shows a structure of the data frame
```

We can look at the particular variable in the data frame by using the dataframe\$variable:

```{r}
friends$names
```

We can also specify the specific subsets of the data frame.\
dataframe[row,column]

```{r}
friends[4,2]
```

Leaving either position blank gives us all the columns for that row, or all rows for that column. Let's leave the column position blank

```{r}
friends[1, ] #show me all columns for the first row
```

## Datasets

The data() function shows all available data sets in R:

```{r}
data()
```

View the content of the data set:

```{r}
View(UCBAdmissions)
```

View the data set in a separate windows:

```{r}
utils::View(UCBAdmissions)
```

### Install Tidyverse which contains the starwars data set

Now we install the tidyverse library. We only need to install a library once.

```{r}
library(tidyverse)
```

Then we load this library so we can use its functions

```{r}
library(tidyverse)
```

This library contains a data set called starwars

```{r}
View(starwars)
```

Pipe operator %\>% means "and then", we use it when applying multiple functions to a data set.

```{r}
starwars %>%
  filter(height > 150 & mass < 200) %>% #filter by these criteria 
  mutate(height_in_meters = height/100) %>% #mutate means change. height_in_meters is the name of the new variable
  select(height_in_meters, mass) %>% #what columns we want to display
  arrange(mass) %>% #order the results by mass from lowest to highest value. we can put a minus sign to order then from highest to lowest
  View() #displays this data set
  #plot() #displays a plot. Use either View or plot, can't use both.
  
```
Ways of displaying a data set called msleep:
```{r}
View(msleep) #dwhole data set
glimpse(msleep) #overwiew - names, types of variables, column and row count
head() #displays first 6 rows
```
What kind of a variable is it? \
class(dataframe$variable)
```{r}
class(msleep$name)
```
Number of variables (technically number of columns)
```{r}
length(msleep)
```
How many rows (observations) in a particular variable (column)?\
length(dataset$variable)
```{r}
length(msleep$name)
```
Show all variable names:
```{r}
names(msleep)
#also possible to use colnames(msleep)
```
Show all unique variable names (displays all names that are at least once in the data set)
```{r}
unique(msleep$order)
```
Show all rows where there is at least one missing value:
```{r}
missing <- !complete.cases(msleep) 
#create a new object called missing
#the ! does the opposite, which means it displays the data that are NOT complete cases
```
Now let's display it:
```{r}
msleep[missing, ] #
```

## Clean
```{r}
starwars %>%
select(name, height, mass) #show me only these variables (columns)
```
Or we can use numbers of the columns:
```{r}
starwars %>%
  select(1:4) #show me first 4 variables
```
Show me all variables whose names end with color:
```{r}
starwars %>%
  select(ends_with("color"))
```

Changing a variable name:
```{r}
#rename("new_name" = "original name")

starwars %>%
  rename("characters" = "name") %>% #rename name to characters
  head() #and show me first 6 rows
```
Changing a variable type:

```{r}
#let's look at what variable is hair_color
class(starwars$hair_color)
```

We can change color variable to a factor variable. Factors are variables with a limited number of different values, often refered to as categorical variables. There only a limited number of colors, so this it makes sense to set this variable to factor:
```{r}
starwars$hair_color <- as.factor(starwars$hair_color)
#hair_color already exists, so we will override it
```
Check if this variable is a factor:
```{r}
class(starwars$hair_color)
```
Let's change it back to character by using mutate:
```{r}
starwars %>%
  mutate(hair_color = as.character(hair_color)) %>%
  glimpse() #check
```

```{r}
starwars %>%
  select(mass, sex) %>% #select these variables/columns
  filter(mass < 55 & sex == "male") #for filtering we use ==
```
Recode data
```{r}
starwars %>%
  select(sex) #this line is not necessary for this example
mutate(sex = recode(sex, "male" = "man", "female" = "woman"))
#we can create or write over existing variable, like here
```

This will create a new variable recoded_sex:
```{r}
starwars %>%
mutate(recoded_sex = recode(sex, "male" = "man", "female" = "woman"))
```

### Dealing with missing data
Check:
```{r}
mean(starwars$height)
#can't calculate mean because there are missing data
```
Now we remove the rows with missing data for this variable using na.rm = TRUE
```{r}
mean(starwars$height, na.rm = TRUE) #remove NA values
```
Show only distinct values (don't show duplicates)
```{r}
friends %>% distinct()
```
### Conditional change
Let's categorize height:
```{r}
starwars %>%
  mutate(height_m = height/100) %>% #create new variable height_m
  select(name, height, height_m) %>%
  mutate(tallness = #create new variable tallness
           if_else(height_m < 1,"short","tall"))
         #if height_m < 1, then create this value called short, else create a value called tall
```


